Sentiment classification in online political discussion -- comparison of two methods

### 2014-03-20 manuscript.tex
added ref to dijkstra
changed opinion mining to sentiment analysis in whole document

@@

\begin{center}
  \textbf{\Large{Classifying opinions in online political discussion -- comparison of two methods}}

\vspace{5mm}

\end{center}

\begin{abstract}
Bla bla
\end{abstract}

@@

Social media and online discussions contain a wealth of information which can make us able to understand the extent of hate speech on the Internet. However, it turns out that academia is lacking research on social media and online radicalization \cite{s1}. Opinion mining is the discipline of automatically extracting opinions from a text material and may be one important tool in the understanding of online radicalization. Opinion mining has mostly been used to analyze opinions in comments and reviews about commercial products, but there are also examples of opinion mining towards political tweets and discussions, see e.g. Tumasjan et al. \shortcite{Tumasjan2010}; Chen et al. \shortcite{Chen10}. Opinion mining towards political discussions is known to be hard since citations, irony and sarcasm is very common \cite{Bing12}.

Opinion classification is perhaps the most studied topic within opinion mining. It aims to classify a set of text as either positive or negative and sometimes also neutral. There are mainly two approaches, one based on machine learning and one based on based on using a list of words with given sentiment scores (lexicon-based approach). One simple lexicon-based approach is to count the number of words with positive and negative sentiment in the document as suggested by Hu and Liu \shortcite{Hu04}. One may classify the opinion of larger documents like movie or product reviews or smaller documents like tweets, comments or sentences. %% See Liu \shortcite{Bing12}, chapters three to five and references therein for the description of several opinion classification methods.

@@

\section{Opinion mining methods}
\label{sec:om}

In this Section we present two methods to classify sentences as either positive, neutral or negative towards a keyword. Both methods follow the same general algorithm presented below which is inspired by Ding et al. \shortcite{Ding08} and is based on a list of sentiment words each associated with a sentiment score representing the polarity and strength of the sentiment word (sentiment lexicon). Both keywords, sentiment words and sentiment shifters can in general appear several times in a sentence. Sentiment shifters are words that potentially shift the sentiment of a sentence from positive to negative or negative to positive. E.g. ``not happy'' have the opposite polarity than just ``happy''. Let $kw_i, i \in \{1,2,\ldots,I\}$ represent appearance number $i$ of the keyword in the sentence. Further let $sw_{j}, j \in \{1,2,\ldots,J\}$ be appearance number $j$ of a sentiment word in the sentence. Finally let $\textbf{ss} = (ss_1, ss_2, \ldots, ss_K)$ repressent the sentiment shifters in the sentence. We compute a sentiment score, $S$, for the sentence as follows
\begin{equation}
  \label{eq:1}
  S = \frac{1}{I}\sum_{i=1}^{I} \sum_{j=1}^{J} \mathbf{imp}(kw_i, sw_{j})\mathbf{shift}(sw_{j}, \mathbf{ss})    
\end{equation}
where the function \textbf{imp} computes the importance of the sentiment word $sw_{j}$ on the keyword appearance $kw_i$. This will be computed in different ways as described below. Further, the function $\mathbf{shift}(sw_{j}, \mathbf{ss})$ computes whether the sentiment of $sw_{j}$ should be shifted based on all the sentiment shifters in the sentence. It returns $-1$ (sentiment shift) if some of the sentiment shifters is within $d_{p}$ words in front or $d_{n}$ words behind $sw_{j}$, respectively. Else the function, returns $1$ (no sentiment shift). We classify the opinion towards the keyword to be positive, neutral or negative if $S >= t_p$,  $t_p > S > t_n$ and  $S <= t_n$, respectively. The parameters $d_p, d_n, t_p$ and $t_n$ is tuned using a training set.

@@

\paragraph{Dependency distance} One way of expressing the syntactic environment of a target word with respect to a sentiment word is to determine its distance in the dependency graph. We therefore define a distance function $\mathbf{depdist}(kw_i, sw_{j})$ which returns the number of nodes in the shortest dependency path from the target word to the sentiment word in the dependency.

@@

We have chosen to give subject-predicate paths the highest possible score, 3. Firstly, the combination of a subject and a predicate will result in a proposition, a statement which is evaluated as true or false \cite{Heim98}. We expect that a proposition typically will represent the opinion of the speaker, although e.g. irony and certain kinds of embedding can shift the truth evaluation in some cases \cite{Hooper73}. Secondly, if the predicate represents an event brought about by an intentional agent, such as \emph{jubler}, 'exult', in (\ref{verbpred}), the subject will typically represent that agent. If the predicate has a positive or negative sentiment, we expect that this sentiment is directed towards this intentional agent.

@@

We compare the classification performance of a set of $M$ different methods, denoted as $\Pi_1, \Pi_2, \ldots, \Pi_M$, using random effect logistic regression. Let the stochastic variable $Y_{tm} \in \{0,1\}$ represents whether method $\Pi_m, \, m \in \{1,2,\ldots,M\}$ classified the correct opinion to sentence number $t \in \{1,2,\ldots,T\}$, where $T$ is the number of sentences in the test set. We let $Y_{tm}$ be the dependent variable of the regression model. The different methods $\Pi_1, \Pi_2, \ldots, \Pi_M$ is included as a categorical independent variable in the regression model. We also assume that classification performance of the different methods depends on the sentence to be classified, thus the sentence number is included as a random effect. The regression model is formulated as
\begin{align}
  \label{eq:5}
  \begin{split}
    P(Y_{tm} = 1) &= \text{logit}(\mu + \Pi_m + \epsilon_t + \epsilon_{tm}) \\
    P(Y_{tm} = 0) &= 1 - P(Y_{tm} = 1)
  \end{split}
 \end{align}
where $\epsilon_t$ is the sentence number random effect and $\epsilon_{tm}$ is additional random noise. Fitting the model to the observed classification performance of the different methods we are able to see if the probability of classifying correctly significantly vary between the methods.

@@

We did not find any suitable annotated text material related to political discussions and therefore created our own. We manually selected 46 debate articles from the Norwegian online newspapers \textit{NRK Ytring, Dagbladet, Aftenposten, VG and Bergens Tidene}. To each debate article there were attached a discussion thread where readers could express their opinions and feelings towards the content of the debate article. All the text from the debate articles and the subsequent discussions were collected using text scraping \cite{Hammer13}. The debate articles were related to religion and immigration and we wanted to classify the opinion towards all words with stem: \textit{islam, muslim, quran, allah, muhammed, imam and mosque}. These are topics that typically create a lot of active discussions and disagreements.

We automatically divided the material into sentences and all sentences containing at least one keyword and one sentiment word were kept for further analysis. If a sentence contained more than one keyword, e.g. both Islam and Quran, the sentence were repeated one time for each keyword in the final text material. We could then classify the opinion towards each of the keywords in the sentence consecutively. To assure that we do not underestimate the uncertainty in the statistical analysis, we see each repetition of the sentence as the same sentence with respect to the sentence random effect in the regression model in Section \ref{sec:sa}

Each sentence was manually annotated as to whether the commenter was positive, negative or neutral towards the keyword in the sentence. Each sentence was evaluated individually. The sentences were annotated based on real-world knowledge, e.g. a sentence like ``Muhammed is like Hitler'' would be annotated as a negative opinion towards Muhammed. Further, if a commenter presented a negative fact about the keyword, the sentence would be denoted as negative.

@@

Finally the material was divided into two parts where the first half of the debate articles with subsequent discussions make up the training set and the rest constitutes a held-out test set. In the development of the important dependency paths, only the training set was used. After the division, the training and test set consisted of a total of 382 and 308 sentences, respectively. Table \ref{tab:1} summarizes the opinions in the sentences
\begin{table}
  \caption{Manual annotation of training and test set.}
  \centering
  \begin{tabular}{lcccc}
             & Negative & Neutral & Positive \\\hline
    Training & 174 (46\%) & 162 (42\%) & 46 (12\%)\\
    Test     & 102 (33\%)& 182 (59\%) & 24 (8\%)
  \end{tabular}
  \label{tab:1}
\end{table}

@@

Several Norwegian sentiment shifters were considered but only the basic shifter 'not' improved the opinion classification and therefore only this word were used in the method.

@@

Classifying opinions in political discussions is hard because of the frequent use of irony, sarcasm and citations. In this paper we have compared the use of word distance between keyword and sentiment word against metrics related to parsed sentence information. Our results show that using dependency tree distances or important dependency paths, improves the classification performance compared to using word distance.

Manually selecting important dependency paths for the aim of opinion mining is a hard task. A natural further step of our analysis is to expand the training and test material and use machine learning to see if there exists dependency paths that improve results compared to using dependency distance.

>>>>

\title{Sentiment classification in online political discussion -- comparison of two methods}

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain} \\}

\date{}

@@

%% \begin{center}
%%   \textbf{\Large{Classifying opinions in online political discussion -- comparison of two methods}}

%% \vspace{5mm}

%% \end{center}
\maketitle
\begin{abstract}
Bla bla
\end{abstract}

@@

Social media and online discussions contain a wealth of information which can make us able to understand the extent of hate speech on the Internet. However, it turns out that academia is lacking research on social media and online radicalization \cite{s1}. Sentiment analysis (SA) is the discipline of automatically determining sentiment in text material and may be one important tool in the understanding of online radicalization. SA has mostly been used to analyze opinions in comments and reviews about commercial products, but there are also examples of SA towards political tweets and discussions, see e.g. Tumasjan et al. \shortcite{Tumasjan2010}; Chen et al. \shortcite{Chen10}. SA of political discussions is known to be a difficult task since citations, irony and sarcasm is very common \cite{Bing12}.

Sentiment classification aims to classify a document or sentence as either positive or negative and sometimes also neutral. There are mainly two approaches, one based on machine learning and one based on based on using a list of words with given sentiment scores (lexicon-based approach). One simple lexicon-based approach is to count the number of words with positive and negative sentiment in the document as suggested by Hu and Liu \shortcite{Hu04}. One may classify the opinion of larger documents like movie or product reviews or smaller documents like tweets, comments or sentences. %% See Liu \shortcite{Bing12}, chapters three to five and references therein for the description of several opinion classification methods.

@@

\section{Proposed SA methods}
\label{sec:om}

In this Section we present two methods to classify sentences as either positive, neutral or negative towards a keyword. Both methods follow the same general algorithm presented below which is inspired by Ding et al. \shortcite{Ding08} and is based on a list of sentiment words each associated with a sentiment score representing the polarity and strength of the sentiment word (sentiment lexicon). Both keywords, sentiment words and sentiment shifters can in general appear several times in a sentence. Sentiment shifters are words that potentially shift the sentiment of a sentence from positive to negative or negative to positive. E.g. ``not happy'' have the opposite polarity than just ``happy''. Let $kw_i, i \in \{1,2,\ldots,I\}$ represent appearance number $i$ of the keyword in the sentence. Further let $sw_{j}, j \in \{1,2,\ldots,J\}$ be appearance number $j$ of a sentiment word in the sentence. Finally let $\textbf{ss} = (ss_1, ss_2, \ldots, ss_K)$ repressent the sentiment shifters in the sentence. We compute a sentiment score, $S$, for the sentence as follows
\begin{equation}
  \label{eq:1}
  S = \frac{1}{I}\sum_{i=1}^{I} \sum_{j=1}^{J} \mathbf{imp}(kw_i, sw_{j})\mathbf{shift}(sw_{j}, \mathbf{ss})    
\end{equation}
where the function \textbf{imp} computes the importance of the sentiment word $sw_{j}$ on the keyword appearance $kw_i$. This will be computed in different ways as described below. Further, the function $\mathbf{shift}(sw_{j}, \mathbf{ss})$ computes whether the sentiment of $sw_{j}$ should be shifted based on all the sentiment shifters in the sentence. It returns $-1$ (sentiment shift) if some of the sentiment shifters is within $d_{p}$ words in front or $d_{n}$ words behind $sw_{j}$, respectively. Else the function, returns $1$ (no sentiment shift). We classify the sentiment towards the keyword to be positive, neutral or negative if $S >= t_p$,  $t_p > S > t_n$ and  $S <= t_n$, respectively. The parameters $d_p, d_n, t_p$ and $t_n$ is tuned using a training set.

@@

\paragraph{Dependency distance} One way of expressing the syntactic environment of a target word with respect to a sentiment word is to determine its distance in the dependency graph. We therefore define a distance function $\mathbf{depdist}(kw_i, sw_{j})$ which returns the number of nodes in the shortest dependency path from the target word to the sentiment word in the dependency. The shortest path is determined using Dijkstra's shortest path algorithm \cite{Dij:59}.

@@

We have chosen to give subject-predicate paths the highest possible score, 3. Firstly, the combination of a subject and a predicate will result in a proposition, a statement which is evaluated as true or false \cite{Heim98}. We expect that a proposition typically will represent the sentiment of the speaker, although e.g. irony and certain kinds of embedding can shift the truth evaluation in some cases \cite{Hooper73}. Secondly, if the predicate represents an event brought about by an intentional agent, such as \emph{jubler}, 'exult', in (\ref{verbpred}), the subject will typically represent that agent. If the predicate has a positive or negative sentiment, we expect that this sentiment is directed towards this intentional agent.

@@

We compare the classification performance of a set of $M$ different methods, denoted as $\Pi_1, \Pi_2, \ldots, \Pi_M$, using random effect logistic regression. Let the stochastic variable $Y_{tm} \in \{0,1\}$ represents whether method $\Pi_m, \, m \in \{1,2,\ldots,M\}$ classified the correct sentiment to sentence number $t \in \{1,2,\ldots,T\}$, where $T$ is the number of sentences in the test set. We let $Y_{tm}$ be the dependent variable of the regression model. The different methods $\Pi_1, \Pi_2, \ldots, \Pi_M$ is included as a categorical independent variable in the regression model. We also assume that classification performance of the different methods depends on the sentence to be classified, thus the sentence number is included as a random effect. The regression model is formulated as
\begin{align}
  \label{eq:5}
  \begin{split}
    P(Y_{tm} = 1) &= \text{logit}(\mu + \Pi_m + \epsilon_t + \epsilon_{tm}) \\
    P(Y_{tm} = 0) &= 1 - P(Y_{tm} = 1)
  \end{split}
 \end{align}
where $\epsilon_t$ is the sentence number random effect and $\epsilon_{tm}$ is additional random noise. Fitting the model to the observed classification performance of the different methods we are able to see if the probability of classifying correctly significantly vary between the methods.

@@

We did not find any suitable annotated text material related to political discussions and therefore created our own. We manually selected 46 debate articles from the Norwegian online newspapers \textit{NRK Ytring, Dagbladet, Aftenposten, VG and Bergens Tidene}. To each debate article there were attached a discussion thread where readers could express their opinions and feelings towards the content of the debate article. All the text from the debate articles and the subsequent discussions were collected using text scraping \cite{Hammer13}. The debate articles were related to religion and immigration and we wanted to classify the sentiment towards all words with stem: \textit{islam, muslim, quran, allah, muhammed, imam and mosque}. These are topics that typically create a lot of active discussions and disagreements.

We automatically divided the material into sentences and all sentences containing at least one keyword and one sentiment word were kept for further analysis. If a sentence contained more than one keyword, e.g. both Islam and Quran, the sentence were repeated one time for each keyword in the final text material. We could then classify the sentiment towards each of the keywords in the sentence consecutively. To assure that we do not underestimate the uncertainty in the statistical analysis, we see each repetition of the sentence as the same sentence with respect to the sentence random effect in the regression model in Section \ref{sec:sa}

Each sentence was manually annotated as to whether the commenter was positive, negative or neutral towards the keyword in the sentence. Each sentence was evaluated individually. The sentences were annotated based on real-world knowledge, e.g. a sentence like ``Muhammed is like Hitler'' would be annotated as a negative sentiment towards Muhammed. Further, if a commenter presented a negative fact about the keyword, the sentence would be denoted as negative.

@@

Finally the material was divided into two parts where the first half of the debate articles with subsequent discussions make up the training set and the rest constitutes a held-out test set. In the development of the important dependency paths, only the training set was used. After the division, the training and test set consisted of a total of 382 and 308 sentences, respectively. Table \ref{tab:1} summarizes the sentiments in the sentences.
\begin{table}
  \caption{Manual annotation of training and test set.}
  \centering
  \begin{tabular}{lcccc}
             & Negative & Neutral & Positive \\\hline
    Training & 174 (46\%) & 162 (42\%) & 46 (12\%)\\
    Test     & 102 (33\%)& 182 (59\%) & 24 (8\%)
  \end{tabular}
  \label{tab:1}
\end{table}

@@

Several Norwegian sentiment shifters were considered but only the basic shifter 'not' improved the sentiment classification and therefore only this word were used in the method.

@@

Classifying sentiment in political discussions is hard because of the frequent use of irony, sarcasm and citations. In this paper we have compared the use of word distance between keyword and sentiment word against metrics related to parsed sentence information. Our results show that using dependency tree distances or important dependency paths, improves the classification performance compared to using word distance.

Manually selecting important dependency paths for the aim of sentiment analysis is a hard task. A natural further step of our analysis is to expand the training and test material and use machine learning to see if there exists dependency paths that improve results compared to using dependency distance.
